---
title: "Deaths and Property Damage Due to Storm Events"
output: html_document
---

## Synopsis
Storms are dangerous events that can cause deaths and property damage. This
document examines the top five lethal and damaging storm events since 2000.

## Data Processing
The three R libraries used for this report are lubridate, dplyr and chorolplethr.
```{r message=FALSE}
  library(lubridate)
  library(dplyr)
  library(choroplethr)
```
The first step in data processing is to download the data and load it into a
data frame.
```{r cache=TRUE}
  event.data.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  temp <- tempfile()
  download.file(event.data.url, temp, method = "curl")
  
  event.data <- read.csv(temp)
  unlink(temp)
```
The downloaded data contains storm events since 1950, however for this report
we are only analyzing events since 2000.
```{r}
  events.since.2000 <- event.data %>%
    filter(STATE %in% state.abb) %>%
    mutate(date = mdy_hms(BGN_DATE)) %>%
    filter(date > mdy("1/1/2000"))
```

We then summarize fatalities and property damage by event type.
```{r}
  deaths.by.event.type <- events.since.2000 %>%
    group_by(EVTYPE) %>%
    summarize(FATALITIES = sum(FATALITIES)) %>%
    arrange(desc(FATALITIES))
  
  exp.to.multiplier = c(1000, 1000000, 1000000000)
  names(exp.to.multiplier) <- c('K', 'M', 'B')
  
  property.damage.by.event.type <- events.since.2000 %>%
    group_by(EVTYPE) %>%
    mutate(PROPDMGEXP.AS.CHAR = as.character(PROPDMGEXP),
           PROPDMGMULTIPLIER = ifelse(
             is.na(exp.to.multiplier[PROPDMGEXP.AS.CHAR]),
             1 ,
             exp.to.multiplier[PROPDMGEXP.AS.CHAR]),
           TRUEPROPDMG = PROPDMG * PROPDMGMULTIPLIER) %>%
    summarize(PROPDMG = sum(TRUEPROPDMG)) %>%
    arrange(desc(PROPDMG))
```

We also calculate the total fatalities by state.
```{r}
  deaths.by.state <- events.since.2000 %>%
    group_by(STATE) %>%
    summarize(FATALITIES = sum(FATALITIES))
```
## Results
As you can see in the first figure the five most deadly storm events are .

```{r}
  top.5.deaths.by.event.type <- head(deaths.by.event.type, 5)
  pie(
    top.5.deaths.by.event.type$FATALITIES,
    top.5.deaths.by.event.type$EVTYPE,
    main = "Top 5 Total Fatalities by Event Type",
    xlab = paste("Total deaths:", sum(top.5.deaths.by.event.type$FATALITIES)))
```

It is also interesting to look at the total deaths by state
```{r warning=FALSE}
  names.of.states <- tolower(state.name[match(deaths.by.state$STATE, state.abb)])
  state_choropleth(
    data_frame(region = names.of.states, value = deaths.by.state$FATALITIES),
    num_colors = 1)
```

The next figure shows a breakdown of the top five most property damaging storm
events
```{r}
  top.5.prop.damage.by.event.type <- head(property.damage.by.event.type, 5)
  total.dmg.for.top5 <- sum(top.5.prop.damage.by.event.type$PROPDMG)
  prop.dmg.as.currency <- paste("$",format(total.dmg.for.top5, big.mark=","),sep="")
  pie(
    top.5.prop.damage.by.event.type$PROPDMG,
    top.5.prop.damage.by.event.type$EVTYPE,
    main = "Top 5 Total Property Damage by Event Type",
    xlab = paste("Total property damage for top 5 property damage events:", prop.dmg.as.currency))
```
